On Optimality of Finite Element Integration

### 2015-11-09 TOMS_Paper/assembly.tex
Hopefully, this clarifies sharing elimination for L_i

\begin{abstract}
We tackle the problem of automatically generating optimal finite element integration routines given a high level specification of arbitrary multilinear forms. Optimality is defined in terms of floating point operations given a memory bound. We provide an approach to explore the transformation space of loops and expressions typical of integration routines and discuss the conditions for optimality. A theoretical analysis and extensive experimentation, which shows systematic performance improvements over a number of alternative code generation systems, validate the approach.
\end{abstract}

@@

To contextualize, consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}. The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a free order loop $L_e$ (over elements in the mesh), a reduction loop $L_i$ (performing numerical integration), and a multilinear loop nest $[L_j, L_k]$ (over test and trial functions). In the body of $L_k$, one or more statements evaluate the local tensor for the element $e$. An expression implements a set of monomials. The expression shown is in \textit{normal form}: $m$ is the (problem-dependent) number of ``terms'', $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry. Given the rigid structure of monomials (summarized in Section~\ref{sec:monomials}), we observe that expressions from any problem can be reduced to this form.

\begin{figure}[h]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest.code}
\end{CenteredBox}\caption{The loop nest implementing a generic bilinear form.}\label{code:loopnest}\end{figure}

We recall that our aim is a strategy to synthesize optimal implementations of loop nests and expressions of this particular form. In particular, we characterize optimality as follows.

@@

\subsubsection{Reduction and free order loops}
\label{sec:se-rln}
Integration- and geometry-dependent expressions should logically be placed, respectively, within the reduction and free order loops. The factorization of $\sigma_{ei}$, therefore, also plays a key role for optimal sharing elimination. This is particularly important in complex forms, characterized by several coefficients, derivatives of coefficients, and extensive use of differential operators. We distinguish two cases:
\begin{itemize}
\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus \lbrace f_i \rbrace$. 
\item There is no common symbol among all terms. There still can be two terms $w_1$ and $w_2$ such that $f_i \in \sigma_{ei}^{w1}, \sigma_{ei}^{w2}$, but factorizing $f_i$ would only increase the operation count since sharing elimination opportunities across test and trial functions would be lost. On the other hand, sharing elimination must occur \textit{within} each term $\sigma_{ei}^{w}$: factorization of (derivatives of) coefficients promotes reduction-loop invariant code motion. 
\end{itemize}

\subsubsection{Observations and underlying assumptions}
\label{sec:se-assumptions}
We observe that there are almost never two ore more conflicting factorizations among which to choose for minimizing the operation count. The idea is that, by choosing a factorization that places a sub-expression within the loop that it logically belongs to, a reduction in operation count is generally obtained. There only are two pathological circumstances in which such a reduction may not be obtained: 
\begin{itemize}
\item As explained in Section~\ref{sec:se-opln}, having both test and trial functions in the same function space exposes sub-expressions over different loops, which introduces a choice between two alternative transformations.
\item The trip count of a loop is extremely small (typically it needs be 1 for the following to happen). This decreases or nullifies the impact of code motion -- exposed through factorization, as described in Sections~\ref{sec:se-opln} and~\ref{sec:se-rln}. For instance, the reader can verify that expanding the expression $(a f_i + b g_i) (c f_i + d g_i)$ within a loop $L_i$\footnote{The example was excerpted from the multiplication of the Jacobian matrix for the change of coordinates with the derivative of a coefficient} and factorizing $L_i$-dependent symbols leads to optimality if and only if the size of $L_i$ is strictly greater than 1.
\end{itemize}

\subsubsection{Algorithm}
Algorithm~\ref{algo:opt} details the steps for sharing elimination assuming as input an abstract syntax tree representation of the loop nest. Neglecting the pathological scenarios described in the previous section, the algorithm performs all necessary transformations and checks to reach optimality. A constructive proof of optimality can be derived taking into account both the algorithm itself and the observations in Sections~\ref{sec:se-opln},~\ref{sec:se-rln} and~\ref{sec:se-assumptions}.

\begin{Algo}[Sharing elimination]
\label{algo:opt}
\normalfont
Consider a finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$. The algorithm starts with reducing the input expression to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$. The sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ are then ``flattened'' (by multiplication expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree.

We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms. The choice of starting with $L_j$ or $L_k$ depends on two parameters: the number of unique terms, $n_{L}$, and the trip count, $tc_{L}$, of a loop $L$. It is fairly simple to see (and easily demonstrable by contradiction) that taking first $L$ such that $n_L \cdot tc_L$ is smallest leads to the largest operation count reduction.

Assume, without loss of generality, that we start with $L_j$. Thanks to linearity, each factored product only has one symbol depending on $L_j$, and this symbol is now unique in the expression tree. The other sub-expression, independent of $L_j$, is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided. Note that multilinearity ensures deterministic factorization, while perfectness ensures correctness of hoisting. We repeat the same process with $L_k$.

For each of the expressions lifted at the level of $L_i$, we factorize $L_i$-dependent terms such that $L_e$-dependent sub-expressions can be hoisted. Every time a sub-expression is hoisted, we search and replace all common sub-expressions. This last step is probably useless for the actual operation count, since we expect any general-purpose compiler to be able to perform aggressive common sub-expression elimination\footnote{By inspection of assembly code, we have verified that this is actually the case with the Intel compiler v15.2.}, but it helps making the generated code much more readable.
\end{Algo}

\subsection{Pre-evaluation}
\label{sec:pre-evaluation}
Sharing elimination reduces the operation count by applying three operators: factorization, code motion and common sub-expression elimination. A fourth operator impacting the operation count is \textit{reduction pre-evaluation}. In this section, we discuss its role and explain that there exists a single point in the whole transformation space (i.e., a specific factorization) that makes applying this operator legal.

@@

With $\iota$ in place, it remains to find the total number of terms, $\rho$, that will be affected by the increase factor. Then we have $\theta_{pre} = \rho \cdot \iota$. Consider again the integration for the mass matrix in Figure~\ref{code:increase_factor}. This form is characterized by the dot product of test and trial functions (\texttt{$b_{ij}$*$b_{ik}$}), so here, trivially, $\rho = 1$. In general, $\rho$ depends on both form and discretization. As another example, take a standard Poisson equation using P1 Lagrange elements over triangles: we have $\rho = 3$ since there are in total four spatial derivatives and factorizing the expression as explained in Algorithm~\ref{algo:pre-evaluation} leads to 3 integration-dependent terms. We determine $\rho$ by executing parts of Algorithm~\ref{algo:pre-evaluation}, which obviously excludes the reduction pre-evaluation.

>>>>

\begin{abstract}
We tackle the problem of automatically generating optimal finite element integration routines given a high level specification of arbitrary multilinear forms. Optimality is defined in terms of floating point operations given a memory bound. We provide an approach to explore the transformation space of a generic integration routine and discuss the conditions for optimality. A theoretical analysis and extensive experimentation, which shows systematic performance improvements over a number of alternative code generation systems, validate the approach.
\end{abstract}

@@

To contextualize, consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}. The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a free order loop $L_e$ (over elements in the mesh), a reduction loop $L_i$ (performing numerical integration), and a multilinear loop nest $[L_j, L_k]$ (over test and trial functions). In the body of $L_k$, one or more statements evaluate the local tensor for the element $e$. An expression (right hand side of a statement) implements a set of monomials. The expression shown is in \textit{normal form}: $m$ is the (problem-dependent) number of ``terms'', $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry. We do not pose particular restrictions on function spaces (e.g., scalar- or vector-valued), coefficients (e.g., linear or non-linear), differential and matrix operators, so $\sigma_{ei}$ can be arbitrarily complex. We observe that translating a monomial in matrix notation into code naturally results in this normal form; this applies to any problem, given the rigid structure of a monomial (summarized in Section~\ref{sec:monomials}). Henceforth, without loss of generality and for ease of treatise, we assume finite element integration loop nests characterized by expressions in normal form.

\begin{figure}\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest.code}
\end{CenteredBox}\caption{The loop nest implementing a generic bilinear form.}\label{code:loopnest}\end{figure}

Our aim is a strategy for reducing finite element integration loop nests to \textit{optimal form}. In particular, we characterize optimality as follows.

@@

\subsubsection{Reduction and free order loops}
\label{sec:se-rln}
The arithmetic complexity of $\sigma_{ei}$ varies with the differential and vector operators employed in a monomial, becoming critical in complicated formulations (characterized, for instance, by several, possibly non-linear coefficients). We distinguish the following cases.
\begin{enumerate}
\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus \lbrace f_i \rbrace$. This is the simplest scenario since factorizing $f_i$ has no side effects.
\item There is no common symbol among all terms, although there still can be two terms $w_1$ and $w_2$ such that $f_i \in \sigma_{ei}^{w1}, \sigma_{ei}^{w2}$. Factorizing $f_i$ would only increase the operation count since sharing elimination opportunities across test and trial functions would be lost (i.e., code motion opportunities are destroyed).
\item Within a term $\sigma_{ei}^{w}$, sharing elimination could still be performed by (i) factorizing integration-dependent symbols and (ii) hoisting geometry-dependent sub-expressions. This is tempting, but not necessarily the optimal choice depending on two factors: the trip count of $L_i$ and the arithmetic complexity of $\sigma_{ei}^{w}$. We provide an example. Consider the sub-expression $(a f_i + b g_i) (c f_i + d g_i)$ in $\sigma_{ei}^{w}$\footnote{The example was excerpted from the multiplication of the Jacobian matrix for the change of coordinates with the derivative of a coefficient}. We note that the form $f_i (a + c) + g_i (b + d)$ (followed by hoisting) reduces the operation count if and only if the trip count of $L_i$ is strictly greater than 1. We see that product expansion, which is essential to expose the whole space of factorizations, increases the operation count (we will see that this concept reiterates throughout the article), so a mechanism to assess the profitability of sharing elimination is required. As we detail in Algorithm~\ref{algo:opt}, the solution will consist of using combinatorics.
\end{enumerate}

\subsubsection{Algorithm}
Algorithm~\ref{algo:opt} details the steps for performing sharing elimination on an abstract syntax tree representation of the loop nest. For simplicity, the algorithm neglects the pathological scenario described in Section~\ref{sec:se-opln}, in which the resulting operation count could only be very close to optimality when both test and trial functions are in the same function space. A constructive proof of optimality can be derived taking into account both the algorithm itself and the observations in Sections~\ref{sec:se-opln} and~\ref{sec:se-rln} (in which we have explored the entire factorization space).

\begin{Algo}[Sharing elimination]
\label{algo:opt}
\normalfont
Consider a finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.

The algorithm starts with ``flattening'' the sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ (by product expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree. We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms. The choice of starting with $L_j$ or $L_k$ depends on two parameters: the number of unique terms, $n_{L}$, and the trip count, $tc_{L}$, of a loop $L$. It is fairly simple to see (and easily demonstrable by contradiction) that taking first $L$ such that $n_L \cdot tc_L$ is smallest leads to the largest operation count reduction. Assume, without loss of generality, that we start with $L_j$. Thanks to linearity, each factored product only has one symbol depending on $L_j$, and this symbol is now unique in the expression tree. The other sub-expression, independent of $L_j$, is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided. Note that multilinearity ensures deterministic factorization, while perfectness ensures correctness of hoisting. We repeat the same process with $L_k$.

For each of the expressions $\eta_{ei} = \sum_w \sigma_{ei}^{w}$ lifted at the level of $L_i$, we need understand whether applying a similar process -- that is, factorizing $L_i$-dependent symbols such that $L_e$-dependent sub-expressions can be hoisted -- actually results in reduced operation count (the problem described in Section~\ref{sec:se-rln}, point 3). Consider two sub-expressions $a_{ei}$ and $b_{ei}$ in $\eta_{ei}$, with $n$ and $m$ being the respective number of summands, and with $c$ being the number of summands in common. Given a product $a_{ei} b_{ei}$, the increase in operation count after product expansion is $n \cdot m - \binom c2$. Given a sum $a_{ei} + b_{ei}$, the operation count after factorization of common terms is $n + m - c$. We use these simple formulae, together with the trip count of $L_i$, to predict the impact -- and then decide on the application -- of sharing elimination at the level of $L_i$.

In all this process, whenever a sub-expression is hoisted, we search and replace all common sub-expressions. This is probably useless for the actual operation count, since we expect any general-purpose compiler to be able to perform aggressive common sub-expression elimination\footnote{By inspection of assembly code, we have verified that this is actually the case with the Intel compiler v15.2.}, but it helps making the generated code much more readable.

% matrix notation, fixed normal form, so no need for fancy algors to do CSE
\end{Algo}

\subsection{Pre-evaluation}
\label{sec:pre-evaluation}
Sharing elimination reduces the operation count by applying three operators: factorization, code motion and common sub-expression elimination. A fourth operator impacting the operation count is \textit{reduction pre-evaluation}. In this section, we discuss its role and explain that there exists a single point in the whole transformation space (i.e., a specific factorization) making applying this operator legal.

@@

It remains to find the total number of terms, $\rho$, that will be affected by the increase factor; that is, $\theta_{pre} = \rho \cdot \iota$. Consider again the mass matrix in Figure~\ref{code:increase_factor}. This monomial is characterized by the dot product of test and trial functions (\texttt{$b_{ij}$*$b_{ik}$}), so here, trivially, $\rho = 1$. In general, $\rho$ depends on both form and discretization. As another example, take a standard Poisson equation using P1 Lagrange elements over triangles: we have $\rho = 3$ since there are in total four spatial derivatives and factorizing the expression as explained in Algorithm~\ref{algo:pre-evaluation} leads to 3 integration-dependent terms. We determine $\rho$ by executing parts of Algorithm~\ref{algo:pre-evaluation}, which obviously excludes the reduction pre-evaluation.
